Fixing Vulnerabilities Potentially Hinders Maintainability

### 2021-01-20 response/response.tex

\documentclass[11pt,fleqn]{article}

\usepackage{vmargin}
\setpapersize{A4}
\setmarginsrb{2.5cm}{2.5cm}{2.5cm}{2.5cm}%
{\baselineskip}{2\baselineskip}{baselineskip}{2\baselineskip}
\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt}

\usepackage{caption}
\usepackage{subcaption}

\usepackage{epsfig}
\usepackage{latexsym}
\usepackage{url}
\usepackage{tgheros}
\renewcommand*\familydefault{\sfdefault}

\newcommand{\eline}{\vspace*{.75\baselineskip}}
\newcommand{\Referee}[1]{\eline \noindent {\bf Reviewer comment #1:} \\}
\newcommand{\Us}{\eline \noindent {\bf Response:}\\}
\newcommand{\TBD}{{\bf To Be Done}}
\newcommand{\newreviewer}[1]{\section*{Reviewer #1}\vspace*{-1.05\baselineskip}}
\newcommand{\editor}[1]{\section*{Editor #1}\vspace*{-1.05\baselineskip}}

%\usepackage{tikz}
\usepackage[skins,breakable]{tcolorbox}
\tcbset{textmarker/.style={
%
skin=enhancedmiddle jigsaw,breakable,parbox=false,before skip=-1mm, after skip=0mm,
boxrule=0mm,leftrule=2mm,rightrule=2mm,boxsep=0mm,arc=0mm,outer arc=0mm,
left=2mm,right=2mm,top=2mm,bottom=2mm,toptitle=1mm,bottomtitle=1mm,oversize}}

\newtcolorbox{rcomment}{textmarker,colback=gray!15!white,colframe=gray!80!white}

\newenvironment{revcomment}[1][]
{\Referee{#1}\begin{rcomment}}
{\end{rcomment}}

\newenvironment{reveditor}
{\begin{rcomment}}
{\end{rcomment}}

\usepackage{todo}
\usepackage{hyperref}

%customized
\usepackage{xfrac}

% more lenient line breaking (avoids text protruding into margins)
\sloppy

\title{\vspace*{-2cm}{\bf Authors' Response to the Review of
 EMSE-D-20-00300:\\
 ``Fixing Vulnerabilities Potentially Hinders Maintainability''}}

\author{Sofia Reis, Rui Abreu, Luis Cruz}
\date{}

\begin{document}

\maketitle

\editor{}

\begin{reveditor}
    Overall, I would like to give you a chance to respond to 
    these issues in a revision of your manuscript but I have 
    to make it clear that it does not guarantee a path towards 
    acceptance. Your rebuttal to \#1 will be of critical importance 
    for me judge the validity of the use of the BCH tool and in 
    making a final decision. Methodology is, of course, very 
    important for ESE journal and \#2 \& \#3s comments on methodological 
    details need to be carefully addressed, too. 
\end{reveditor}

\Us We thank the editor and reviewers for their valuable feedback. In the 
following points, we address the concerns raised by the reviewers.

\newreviewer{1}

\begin{revcomment}[1.1]

    This paper investigates the impact of patches to improve 
    security on the maintainability of open-source software. 
    The paper is well written and easy to read. I really 
    appreciated the replication package: complete, and well 
    described. Really a good work!

    However, I found a crucial issue in this paper. The authors 
    adopted as static analysis tool Better Code Hub. Better Code 
    Hub' model includes 10 guidelines that can help the developers to 
    write better code. Unfortunately, these guidelines are not related 
    to maintainability. Avoiding introducing the issues associated to 
    these guidelines do not imply increasing the maintainability of the 
    code, since the tools has no possibility to measure "maintainability".

\end{revcomment}

\Us Thank you for your positive feedback on the paper presentation and replication
package. BCH is not a static analysis tool instead it checks codebases for 
compliance with the ten guidelines. The set of guidelines presented in our paper
is the same published by Software Improvement Group (SIG) in their ebook called
"Building Maintainable Software: Ten Guidelines for Future-Proof Code"
\footnote{Available here: https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}.
According to the authors, all the guidelines were infered from analyzing
hundres of real-world systems and measure maintainability.

We clarified the difference between BCH and static analysis tools in
the Section 1 (Introduction) and Section 3 (Methodology).

\begin{revcomment}[1.2]

    As far as I know, the company that developed Better Code Hub 
    developed also the Delta Maintainability Model [diBiase2019] 
    that aims at measuring the maintainability of a code change 
    and a score and compare change-based maintainability measurements.

    [diBiase2019] M. di Biase and A. Rastogi and M. Bruntink and A. van 
    Deursen. The Delta Maintainability Model: Measuring Maintainability 
    of Fine-Grained Code Changes. IEEE/ACM International Conference 
    on Technical Debt (TechDebt) in 2019.

\end{revcomment}

\Us Thank you for suggesting Marco's paper. Our paper uses a model 
for measuring maintainability that was published previously by one of the co-authors
of this paper [Cruz2019]. The entire methodology and calculations 
were actually validated by Marco.

The model created by Marco focus on measuring maintainability of fine-grained 
changes using 5 of the same guidelines we use. The SIG-MM model in 
Marco's paper is the model behind Better Code Hub.

We added Marco's work to our related work.

[Cruz2019] Luis Cruz, Rui Abreu, John Grundy, Li Li, Xin Xia. 
Do Energy-oriented Changes Hinder Maintainability?. International 
Conference on Software Maintenance and Evolution (ICSME) in 2019.

\begin{revcomment}[1.3]

    Moreover, I am not sure that Better Code Hub is 
    the best static analysis tools to detect security 
    issues. One of the most adopted in security domain 
    is Coverity Scan for examples, but also other tools 
    might be a better choice (e.g. Kiuwan).

\end{revcomment}

\Us Thank your for your comment. In this paper, we do not
try to use static analysis to detect security issues.
We have already a dataset of security patches (vulnerable versions
and safe versions) and our goal is to understand what 
is the impact of those patches on the maintainability
guidelines/metrics. We arguee that these guidelines/metrics
can in the future complement static analysis tools 
by assiting developers with more information on the 
risks associated with their patches.

SonarQube uses 4 of the set of metrics we are using and 
also rates maintainability. In Section X, 
we compare the results between BCH and SonarQube. However, 
we arguee that BCH performs a more complete analysis.

\begin{revcomment}[1.4]

    The authors could have adopted both approaches, Better 
    Code Hub guidelines or another one first and then measure 
    the code maintainability with the Delta Maintainability Model 
    since the goal of the paper is to improve security on the 
    maintainability.

\end{revcomment}

\Us Thank you for your comment. As we explained before, while 
addressing comment 1.2, both models assess maintainability.
Delta Maintainability Model is an alternative to our approach 
that uses only 5 of the set of guidelines our study considers.

\begin{revcomment}[1.5]

    In my opinion, the idea behind this work is very interesting, 
    but the research questions cannot be answered with the tools 
    and metrics selected.

    My recommendation is to select a specific static analysis tool 
    for security issues and to include a maintainability model to 
    evaluate the maintenance.

\end{revcomment}

\Us Thank you for the positive interest on the work. As we 
explained before Better Code Hub calculates maintainability metrics.
Our goal was to measure the impact of security patches on software 
maintainability. For future research, it would be interesting to 
understand if the results of these metrics can actually help 
security engineers assessing the risk of their patches and guide 
them on performing better patches. However, in this work, 
we only focus on assessing software maintainability.

\newreviewer{2}

\begin{revcomment}[2.1]

    Question:  Is your dataset available?

\end{revcomment}

\Us Thank you for your question. The dataset is fully available at 
\texttt{dataset/db\_release\_security\_fixes.csv} in the figshare package
provided in the contributions of the paper: 
\url{https://figshare.com/s/4861207064900dfb3372}. If the paper is 
accepted, the package will be available on GitHub.

\begin{revcomment}[2.2]

    Abstract:
    Throughout the paper, "hypothesize" is probably a better word than 
    "suspect" for sounding more scientific.

    Your results should be more specific than "show evidence of trace-off". 
    Briefly tell us about what metrics you used and what the numerical results 
    indicate.

\end{revcomment}

\Us We thank the reviewer for the suggestions. We addressed
both of the issues in the abstract. For the later issue, we reported 
the overall result for maintainability and results
for the two of the guidelines with more negative impact in software 
maintainability: software complexity and unit size.

\end{document}

### 2021-01-20 response/response.tex

    Your results should be more specific than "show evidence of trace-off". 
    Briefly tell us about what metrics you used and what the numerical results 
    indicate.

\end{revcomment}

\Us We thank the reviewer for the suggestions. We addressed
both of the issues in the abstract. For the later issue, we reported 
the overall result for maintainability and results
for the two of the guidelines with more negative impact in software 
maintainability: software complexity and unit size.

\begin{revcomment}[2.3]

    Intro:
    \begin{itemize}
        \item First sentence - quality is not ONLY related to cost but also to security 
        and safety.
        \item Provide a URL to Software Improvement Group and Better Code Hub.
        \item Page 3, Line 1 - Application Security Verification Standard (ASVS).
        \item Page 3, Line 7 - instead of a "broad number of code metrics" - tell 
        us exactly the number of metrics.
        \item Page 3, Line 13 - "suggest" sounds more scientific than "hint at"
        \item Page 3, Line 24 - "we intend to highlight the need …". Is that the 
        broad goal of your paper?  The goal should be explicitly stated in the 
        abstract and intro.
    \end{itemize}
\end{revcomment}

\Us Thank you for reporting these minor issues. All of them were addressed in 
the new version. In the last point, where the reviewer ask ``Is that the 
broad goal of your paper?'', we want to clarify that the goal of our study
is to show the impact of security patches on sofwtare maintainability and
highlight the need for solutions which are described in the introduction
and end of the abstract.

\begin{revcomment}[2.4]

    Intro:\\
    You should check out this paper: 
    Li, Frank and Paxson, Vern. A large-scale empirical study of security patches.
    ACM SIGSAC Conference on Computer and Communications Security in 2017.

\Us Thank you for bringing our attention to this paper. We integrated
this paper in the Related Work section and used it to support some 
of our claims and findings.

### 2021-01-20 paper/emse20.tex
small fix to abstract
more changes and responses for reviewer 2

\begin{abstract}
Security is a requirement of utmost importance to produce 
high-quality software. However, there is still a considerable amount 
of vulnerabilities being discovered and fixed, almost weekly. We 
hypothesize that developers affect the maintainability of their 
codebases when patching vulnerabilities. This paper evaluates the 
impact of patches to improve security on the maintainability of 
open-source software. Maintainability is measured based on the 
Better Code Hub’s model of 10 guidelines on a dataset including 
1300 security-related commits. Results show evidence of a trade-off 
between security and maintainability in $41.9\%$ of the cases, i.e., developers 
may hinder software maintainability. More than $35\%$ of the security 
patches studied increased software complexity and the percentage 
of LOCs per unit. The implications of our study 
are that changes to codebases while patching vulnerabilities need to 
be performed with extra care; tools for patch risk assessment should 
be integrate into the CI/CD pipeline; computer science curricula 
needs to be updated; and, more secure programming languages are 
necessary.
\keywords{Software Security \and Software Maintenance \and Open-Source Software}
\end{abstract}

@@

Researchers found a correlation between the presence of 
vulnerabilities on software and code 
complexity~\cite{shin2010evaluating,10.1145/1774088.1774504}. 
Security experts claim that complexity hides bugs that may result in 
security 
vulnerabilities~\cite{mcgraw2004software,schneier2006beyond}. In 
practice, an attacker only needs to find one way into the system 
while a defender needs to find and mitigate all the security issues. 
Complex code is difficult to understand, maintain and 
test~\cite{1702388}. Thus, the developer's task gets harder as the 
codebase grows in size and complexity. But the risk can be minimized 
by writing clean and maintainable code.

ISO describes \textit{software maintainability} as ``the degree of 
effectiveness and efficiency with which a software product or system 
can be modified to improve it, correct it or adapt it to changes in 
environment, and in requirements'' on software quality ISO/IEC 
25010~\cite{iso:2011}. Thereby, maintainable security may be 
defined, briefly, as the degree of effectiveness and efficiency with 
which software can be changed to mitigate a security 
vulnerability---corrective maintenance.
% (hopefully, without introducing new ones). 
However, many developers still lack knowledge on the best 
practices to deliver and maintain secure and high-quality 
software~\cite{Pothamsetty:2005:SEL:1107622.1107635,8077802}. In a 
world where zero-day vulnerabilities are constantly emerging, 
mitigation needs to be fast and efficient. Therefore, it 
is important to write maintainable code to support the production of 
more secure software---maintainable code is less complex and, 
consequently, less prone to
vulnerabilities~\cite{shin2010evaluating,10.1145/1774088.1774504})---
and, prevent the introduction of new vulnerabilities.
% Unmaintainable code is hard to test, analyze and re-use. Therefore,
% it is important to write maintainable code to support the production
% of more secure software and prevent
% the introduction of new vulnerabilities in the code bases.
%
Novel tools have been built to automatically detect software 
vulnerabilities (e.g., FindBugs, Infer and more). Despite these 
efforts, improving software security is not a trivial task and 
requires implementing patches that might affect software 
maintainability. Our suspicion is that some of these patches may 
have a negative impact on the software maintainability and, 
possibly, even be the cause of the introduction of new 
vulnerabilities---harming software reliability and introducing 
technical debt. Therefore, in this paper, we present an empirical 
study on the impact of patches of vulnerabilities on software 
maintenance across open-source software.

As ISO does not provide any specific guidelines/formulas to 
calculate maintainability, we resort to Software Improvement Group 
(SIG)'s web-based source code analysis service Better Code Hub (BCH) 
to compute the software compliance with a set of $10$ 
guidelines/metrics to produce quality software based in ISO/IEC 
$25010$~\cite{Visser:2016:OREILLY}. There are other well-known 
standards and models that have been proposed to increase software 
security: Common Criteria~\cite{common:2009} which received
negative criticism regarding the costs associated and poor technical 
evaluation; the OWASP Application Security Verification 
Criteria~\cite{oswap:2009} which is focused only on web 
applications, and a model proposed by Xu et al. ($2013$)
~\cite{6616351} for rating software security (arguably, it was one 
of the first steps taken by SIG to introduce security on their 
maintainability model). Nevertheless, our study uses BCH to provide 
an assessment of maintainability in software for the following 
reasons: BCH integrates a broad number of code metrics; and, 
code metrics were empirically validated in previous 
work~\cite{Bijlsma:2012:FIR:2317098.2317124,8530041,8919169,8785997}.

From a methodological perspective, we leveraged a dataset of $1300$ 
security patches collected from open-source software. We calculate 
software maintainability before and after the patch to measure its 
impact. This empirical study presents evidence that changes applied 
in the codebases to patch vulnerabilities affect code 
maintainability. Results also hint that developers 
should pay different levels of attention to different severity 
levels and classes of weaknesses when patching vulnerabilities. We 
also show that patches in programming languages such as, 
\emph{C/C++}, \emph{Ruby} and \emph{PHP}, may have a more negative 
impact in software maintainability. Little information is known 
about the impact of security patches on software 
maintainability. Developers need to be aware of the impact of their 
changes on software maintainability while patching security 
vulnerabilities. The harm of maintainability can increase the time 
of response of future mitigations or even other maintainability 
tasks. Thus, it is of utmost importance to find means to assist 
mitigation and reduce its risks. With this study, we intend to 
highlight the need for tools to assess the impact of patches on 
software maintainability~\cite{4724577}; the importance of 
integrating maintainable security in computer science curricula; 
and, the demand for better programming languages, designed by 
leveraging security principles~\cite{kurilova2014wyvern,10.1145/2489828.2489830}.

This research performs the following main contributions:
%
\begin{itemize}
	\vspace{-0.3em}
  \item Evidence that supports the trade-off between security and 
  maintainability: developers may be hindering software 
  maintainability while patching vulnerabilities.
	\item An empirical study on the impact of security patches on 
	software maintainability (per guideline, severity, weakness and 
	programming language).
	\item A replication package with all scripts and data created to 
	perform the empirical evaluation, for reproducibility. Available 
	online: \url{https://figshare.com/s/4861207064900dfb3372}.
\end{itemize}
%
This paper is structured as follows: Section~\ref{sec:motivation} 
introduces an example of a security patch of a known vulnerability 
found in the protocol implementation of 
OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
contains open-source implementations of the SSL and TLS cryptographic
protocols. Repository available at 
\url{https://github.com/openssl/openssl} (Accessed on \today{})}; 
Section~\ref{sec:methodology} describes the methodology used to 
answer the research questions; Section~\ref{sec:results} presents 
the results and discusses their implications; 
Section~\ref{sec:implications} elaborates on the implications
developers should consider in the future; Section~\ref{sec:threats} 
enumerates the threats to the validity of this study; 
Section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, 
Section~\ref{sec:conclusions} concludes the main findings and 
elaborates on future work.

\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security 
expertise~\cite{8077802}, code-related security flaws are generally 
detected \textit{after the fact}, i.e., when hackers exploit them. 
Patching these issues is usually done by modifying the codebase. 
However, these patches may have a negative impact on software 
maintenance (e.g., developers might follow the quickest solution to 
fix the problem but not necessarily the most elegant and performant 
one). Thus, we want to understand how security patches impact
software maintainability.

@@

\textit{\textbf{Construct Validity:}} The formula to calculate the maintainability value ($M(v)$) was inferred based 
on the BCH's reports. The high amount of different projects and backgrounds 
may require other maintainability standards. However, BCH does use a 
representative benchmark of closed and open-source software projects to compute 
the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY,Baggen2012}.
Maintainability is computed as the mean of all guidelines. Different software versions 
(vulnerable/fixed) of one vulnerability may have the same overall score
and still be affected by different guidelines. Therefore, we provide
an analysis per guideline and our results are all available on figshare for future reproductions and deeper analysis.

>>>>

\begin{abstract}
Security is a requirement of utmost importance to produce 
high-quality software. However, there is still a considerable amount 
of vulnerabilities being discovered and fixed, almost weekly. We 
hypothesize that developers affect the maintainability of their 
codebases when patching vulnerabilities. This paper evaluates the 
impact of patches to improve security on the maintainability of 
open-source software. Maintainability is measured based on the 
Better Code Hub’s model of 10 guidelines on a dataset including 
1300 security-related commits. Results show evidence of a trade-off 
between security and maintainability in $41.9\%$ of the cases, i.e., developers 
may hinder software maintainability. More than $35\%$ of security 
patches increased software complexity and the percentage 
of LOCs per unit. The implications of our study 
are that changes to codebases while patching vulnerabilities need to 
be performed with extra care; tools for patch risk assessment should 
be integrate into the CI/CD pipeline; computer science curricula 
needs to be updated; and, more secure programming languages are 
necessary.
\keywords{Software Security \and Software Maintenance \and Open-Source Software}
\end{abstract}

@@

Researchers found a correlation between the presence of 
vulnerabilities on software and code complexity~\cite{shin2010evaluating,10.1145/1774088.1774504}. 
Security experts claim that complexity hides bugs that may result in 
security vulnerabilities~\cite{mcgraw2004software,schneier2006beyond}. In 
practice, an attacker only needs to find one way into the system 
while a defender needs to find and mitigate all the security issues. 
Complex code is difficult to understand, maintain and 
test~\cite{1702388}. Thus, the task of a developer gets harder as the 
codebase grows in size and complexity. But the risk can be minimized 
by writing clean and maintainable code.

ISO describes \textit{software maintainability} as ``the degree of 
effectiveness and efficiency with which a software product or system 
can be modified to improve it, correct it or adapt it to changes in 
environment, and in requirements'' on software quality ISO/IEC 
25010~\cite{iso:2011}. Thereby, maintainable security may be 
defined, briefly, as the degree of effectiveness and efficiency with 
which software can be changed to mitigate a security 
vulnerability---corrective maintenance.
% (hopefully, without introducing new ones). 
However, many developers still lack knowledge on the best 
practices to deliver and maintain secure and high-quality 
software~\cite{Pothamsetty:2005:SEL:1107622.1107635,8077802}. In a 
world where zero-day vulnerabilities are constantly emerging, 
mitigation needs to be fast and efficient. Therefore, it 
is important to write maintainable code to support the production of 
more secure software---maintainable code is less complex and, 
consequently, less prone to
vulnerabilities~\cite{shin2010evaluating,10.1145/1774088.1774504})---
and, prevent the introduction of new vulnerabilities. 
% Unmaintainable code is hard to test, analyze and re-use. Therefore,
% it is important to write maintainable code to support the production
% of more secure software and prevent
% the introduction of new vulnerabilities in the code bases.
%
Static analysis tools (SATs) have been built to automatically detect software 
vulnerabilities (e.g., FindBugs, Infer and more). Developers
use those tools to locate the issues in the code. However,
while performing the patches to those issues, SATs are not 
capable of providing information on the quality of the patch.
Improving software security is not a trivial task and 
requires implementing patches that might affect software 
maintainability. Our hypothesis is that some of these patches may 
have a negative impact on the software maintainability and, 
possibly, even be the cause of the introduction of new 
vulnerabilities---harming software reliability and introducing 
technical debt. Research found that $34\%$ of the security patches 
performed introduce new problems and $52\%$ are incomplete and do not 
fully secure systems~\cite{10.1145/3133956.3134072}. Therefore, in this paper, 
we present an empirical study on the impact of patches of 
vulnerabilities on software maintenance across open-source software.
We arguee that tools that assess these type of metrics may
complement SATs with valuable information to understand and
address better the issue.

As ISO does not provide any specific guidelines/formulas to 
calculate maintainability, we resort to Software Improvement Group 
(SIG\footnote{SIG's website: 
\url{https://www.sig.eu/} (Accessed on \today{})})'s web-based source 
code analysis service Better Code Hub (BCH)\footnote{BCH's 
website: \url{https://bettercodehub.com/} (Accessed on \today{})}  
to compute the software compliance with a set of $10$ 
guidelines/metrics to produce quality software based in ISO/IEC 
$25010$~\cite{Visser:2016:OREILLY}. SIG 
has been helping business and technology leaders drive their organizational 
objectives by fundamentally improving the health and security of 
their software applications for more than 20 years. Their 
models are scientically proven~\cite{5609747,6113040,baggen2012} and certified.

There are other well-known 
standards and models that have been proposed to increase software 
security: Common Criteria~\cite{common:2009} which received
negative criticism regarding the costs associated and poor technical 
evaluation; the OWASP Application Security Verification 
Standard (ASVS)~\cite{oswap:2009} which is focused only on web 
applications, and a model proposed by Xu et al. ($2013$)
~\cite{6616351} for rating software security (arguably, it was one 
of the first steps taken by SIG to introduce security on their 
maintainability model). Nevertheless, our study uses BCH to provide 
an assessment of maintainability in software for the following 
reasons: BCH integrates a total of $10$ different code metrics; and, 
code metrics were empirically validated in previous 
work~\cite{Bijlsma:2012:FIR:2317098.2317124,8530041,8919169,8785997}.

From a methodological perspective, we leveraged a dataset of $1300$ 
security patches collected from open-source software. We calculate 
software maintainability before and after the patch to measure its 
impact. This empirical study presents evidence that changes applied 
in the codebases to patch vulnerabilities affect code 
maintainability. Results also suggest that developers 
should pay different levels of attention to different severity 
levels and classes of weaknesses when patching vulnerabilities. We 
also show that patches in programming languages such as, 
\emph{C/C++}, \emph{Ruby} and \emph{PHP}, may have a more negative 
impact in software maintainability. Little information is known 
about the impact of security patches on software 
maintainability. Developers need to be aware of the impact of their 
changes on software maintainability while patching security 
vulnerabilities. The harm of maintainability can increase the time 
of response of future mitigations or even other maintainability 
tasks. Thus, it is of utmost importance to find means to assist 
mitigation and reduce its risks. With this study, we intend to 
highlight the need for tools to assess the impact of patches on 
software maintainability~\cite{4724577}; the importance of 
integrating maintainable security in computer science curricula; 
and, the demand for better programming languages, designed by 
leveraging security principles~\cite{kurilova2014wyvern,10.1145/2489828.2489830}.

This research performs the following main contributions:
%
\begin{itemize}
	\vspace{-0.3em}
  \item Evidence that supports the trade-off between security and 
  maintainability: developers may be hindering software 
  maintainability while patching vulnerabilities.
	\item An empirical study on the impact of security patches on 
	software maintainability (per guideline, severity, weakness and 
	programming language).
	\item A replication package with the scripts and data created to 
	perform the empirical evaluation, for reproducibility. Available 
	online: \url{https://figshare.com/s/4861207064900dfb3372}.
\end{itemize}
%
This paper is structured as follows: Section~\ref{sec:motivation} 
introduces an example of a security patch of a known vulnerability 
found in the protocol implementation of 
OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
contains open-source implementations of the SSL and TLS cryptographic
protocols. Repository available at 
\url{https://github.com/openssl/openssl} (Accessed on \today{})}; 
Section~\ref{sec:methodology} describes the methodology used to 
answer the research questions; Section~\ref{sec:results} presents 
the results and discusses their implications; 
Section~\ref{sec:implications} elaborates on the implications
developers should consider in the future; Section~\ref{sec:threats} 
enumerates the threats to the validity of this study; 
Section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, 
Section~\ref{sec:conclusions} concludes the main findings and 
elaborates on future work.

\section{Motivation and Research Questions}\label{sec:motivation}
%
% Due to time-to-market pressure and the lack of security 
% expertise~\cite{8077802}, code-related security flaws are generally 
% detected \textit{after the fact}, i.e., when hackers exploit them. 
% Patching these issues is usually done by modifying the codebase. 
% However, these patches may have a negative impact on software 
% maintenance (e.g., developers might follow the quickest solution to 
% fix the problem but not necessarily the most elegant and performant 
% one). Thus, we want to understand how security patches impact
% software maintainability.

@@

\textit{\textbf{Construct Validity:}} The formula to calculate the maintainability value ($M(v)$) was inferred based 
on the BCH's reports. The high amount of different projects and backgrounds 
may require other maintainability standards. However, BCH does use a 
representative benchmark of closed and open-source software projects to compute 
the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY,baggen2012}.
Maintainability is computed as the mean of all guidelines. Different software versions 
(vulnerable/fixed) of one vulnerability may have the same overall score
and still be affected by different guidelines. Therefore, we provide
an analysis per guideline and our results are all available on figshare for future reproductions and deeper analysis.
